---
title: "p8105_hw3_jz3902"
author: "Jinghan Zhao"
date: "2024-10-11"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(fig.width = 10, 
                      fig.height = 5,
                      out.width = "90%",
                      message = FALSE)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

### Short description

```{r import_noaa}
library(p8105.datasets)
data("ny_noaa")

ny_noaa
```

There are `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns in the dataset.

```{r missing_noaa}
colMeans(is.na(ny_noaa))
```

We have missing values in `prcp`, `snow`, `snwd`, `tmax` and `tmin` columns. Especially, there over 40% data is missing in `tmax` and `tmin`, which is a rather huge proportion. 

This issue may result in a large amount of observations being ignored during data analysis, making the analysis results incomplete and unable to reflect the true situation.


### Data cleaning

```{r data_cleaning_noaa}
ny_noaa_clean = 
  ny_noaa %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    prcp = prcp / 10,
    tmax = tmax / 10,
    tmin = tmin / 10
  ) %>% 
  relocate(id, date, year, month, day)

ny_noaa_clean %>% 
  count(snow) %>% 
  arrange(desc(n)) %>% 
  head()
```

a) Data cleaning steps:

* Create variables: `year`, `month` and `day`

* Arrange the units of variables: `prcp`(mm), `snow`(mm), `snwd`(mm), `tmax`(degrees C), `tmin`(degrees C)

* Organize variables in a meaningful order

b) For snowfall, the most commonly observed values is 0. Because the dataset covers the whole year, but snow usually occurs in winter, most observations do not have snow.


### Average tmax in Jan and Jul

```{r avg_tmax}
month.labs = c("January", "July")
names(month.labs) = c(1, 7)

ny_noaa_clean %>% 
  filter(month %in% c(1, 7)) %>%
  group_by(id, year, month) %>% 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>% 
  drop_na(avg_tmax) %>% 
  ggplot(aes(x = year, y = avg_tmax, color = id)) +
  geom_point() +
  geom_line() +
  facet_grid(. ~ month, 
             labeller = labeller(month = month.labs)) +
  labs(
    title = "Average Maximum Temperature in Each Station",
    x = "Year",
    y = "Average Max Temp (°C)",
    caption = "The average max temperature in January and in July in each station across years.") +
  theme(legend.position = "none")
```

* Most of the time, the average maximum temperatures at different stations have similar trends: they rise or fall at the same time. 

* According to the graph, outliers occurred in July between 1985-1990. To find out which years they are:

```{r avg_tmax_outlier}
ny_noaa_clean %>% 
  filter(month == 7) %>%
  group_by(id, year, month) %>% 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>% 
  drop_na(avg_tmax) %>% 
  arrange(avg_tmax) %>% 
  head()
```

* Station USC00308962 have extreme low average temperatures (14.0 degrees C) in January, 1988.


### tmax vs tmin + snowfall

```{r tmax_tmin_snowfall}
hex_tmax_tmin = 
  ny_noaa_clean %>% 
  drop_na(tmin, tmax) %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(
    title = "Hexbin Plot: tmax vs tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8, angle = 45))
  

box_snowfall = 
  ny_noaa_clean %>% 
  select(year, snow) %>% 
  filter(snow > 0 & snow < 100) %>% 
  ggplot(aes(x = factor(year), y = snow)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Distribution of Snowfall (between 0-100 mm)",
       x = "Year",
       y = "Snowfall (mm)") +
  theme(axis.text.x = element_text(size = 8,
                                   angle = 45, 
                                   hjust = 1))

hex_tmax_tmin + box_snowfall
```



## Problem 2

### Organizing datasets

```{r mims_tidy}
covar_df = 
  read_csv("HW3_data/nhanes_covar.csv", skip = 4) %>% 
  janitor::clean_names()

accel_df = 
  read_csv("HW3_data/nhanes_accel.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minute",
    values_to = "mims",
    names_prefix = "min"
  )

mims_df = 
  left_join(covar_df, accel_df, by = "seqn") %>% 
  filter(age >= 21) %>% 
  drop_na(sex, age, bmi, education) %>% 
  mutate(sex = factor(sex, 
                      levels = c(1, 2),
                      labels = c("male", "female"),
                      ordered = FALSE
                      ),
         education = factor(education,
                            levels = c(1, 2, 3),
                            labels = c("Less than high school",
                                       "High school equivalent",
                                       "More than high school"),
                            ordered = TRUE
                            )
         )
```

Steps:

* Import the dataset, omit non-data entries

* Pivot `accel_df` longer

* Combine the two datasets, exclude participants: (i) less than 21 years of age; (ii) those with missing demographic data

* Convert numeric variable `sex` and `education` to reasonable factor classes (`sex` is not ordered and `education` is ordered)


### Sex-related distributions

```{r sex_education_table}
mims_df %>% 
  select(sex, education) %>% 
  group_by(sex, education) %>% 
  summarize(n_obs = n()) %>% 
  pivot_wider(
    names_from = education,
    values_from = n_obs
  ) %>% 
  knitr::kable()
```

* For the “less than high school” and “more than high school” categories, the numbers of male and female participants are similar.

* For the "high school equivalent" category, the number of male participants is significantly greater than that of female participants, approximately 1.5 times as many.

```{r age_distribution}
mims_df %>% 
  select(sex, education, age) %>% 
  group_by(sex, education) %>% 
  ggplot(aes(x = education, y = age, fill = sex)) +
  geom_boxplot(alpha = .5) +
  labs(
    title = "Age Distribution",
    x = "Education Category",
    y = "Age",
    fill = "Sex",
    caption = "The age distributions for men and women in each education category.")
```

* For the “less than high school” and “more than high school” categories, the median age and middle 50% range are similar for male and female participants.

* For the "high school equivalent" category, the middle 50% age range for male participants is smaller than female participants. The main age groups of male and female participants are different, which may affect the comparability of data between different genders.


### Total activity vs age

```{r total_activ}
mims_df %>% 
  group_by(seqn, age, sex, education) %>% 
  summarize(total_activ = sum(mims)) %>% 
  ggplot(aes(x = age, y = total_activ, color = sex)) +
  geom_point() +
  facet_grid(. ~education) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Total Accelerometer Activity vs Age",
    x = "Age",
    y = "Total Activity",
    color = "Sex",
    caption = "The total accelerometer activities over the day across different ages, comparing between genders and education levels.") 
```

* Overall trend: with increasing age, total accelerometer activities throughout the day decreases.

* The total activities in the "more than high school" category are more evenly distributed across different ages.

* For different educational levels, the downward trend is more obvious for those with “less than high school” and more gradual for those with “more than high school”. 

* For different genders, the overall trend is similar. In "high school equivalent" category, the difference between male and female participants is significant, and the data fluctuates. Combined with the previous tables and plots, this result requires further analysis due to the differences in the number and age distribution of participants.

### 24-hour activity

```{r 24hour_activ}
mims_df %>% 
  mutate(
    minute = as.numeric(minute),
    hour = minute %/% 60
    ) %>% 
  group_by(seqn, hour, sex, education) %>% 
  summarize(hour_activ = sum(mims)) %>% 
  ggplot(aes(x = hour, y = hour_activ, color = sex)) +
  geom_point(alpha = .5) +
  facet_grid(. ~education) +
  geom_smooth(se = FALSE) +
  labs(
    title = "24-hour Accelerometer Activity",
    x = "Hour",
    y = "Hour Activity",
    color = "Sex",
    caption = "The 24-hour accelerometer activities for men and women in each education category.")
```

* Overall trend: starting from midnight, the accelerometer data is low during the midnight; it starts to rise around 5 am and reaches the maximum value of the day around 10 am; then it starts to decline slowly until it starts to decline rapidly around 8 pm.

* For “less than high school” category, accelerometer activities for male and female are similar.

* For "high school equivalent" category, accelerometer activities for male and female are slightly different: the maximum value of the day is slightly larger for female participants.

* For "more than high school" category, female participants’ accelerometer activity remain relatively balanced during the day (rather than gradually decreasing) and has slightly larger values than male participants. Besides, there are a few outliers with extremely high activity values in this category (hour accelerometer activity > 2500), which are not observed in other two categories.


## Problem 3

