---
title: "p8105_hw3_jz3902"
author: "Jinghan Zhao"
date: "2024-10-11"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
```

## Problem 1

### Short description

```{r import_noaa}
library(p8105.datasets)
data("ny_noaa")

ny_noaa
```

There are `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns in the dataset.

```{r missing_noaa}
colMeans(is.na(ny_noaa))
```

We have missing values in `prcp`, `snow`, `snwd`, `tmax` and `tmin` columns. Especially, there over 40% data is missing in `tmax` and `tmin`, which is a rather huge proportion. 

This issue may result in a large amount of observations being ignored during data analysis, making the analysis results incomplete and unable to reflect the true situation.


### Data cleaning

```{r data_cleaning_noaa}
ny_noaa_clean = 
  ny_noaa %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    prcp = prcp / 10,
    tmax = tmax / 10,
    tmin = tmin / 10
  ) %>% 
  relocate(id, date, year, month, day)

ny_noaa_clean %>% 
  count(snow) %>% 
  arrange(desc(n)) %>% 
  head()
```

a) Data cleaning steps:

* Create variables: `year`, `month` and `day`

* Arrange the units of variables: `prcp`(mm), `snow`(mm), `snwd`(mm), `tmax`(degrees C), `tmin`(degrees C)

* Organize variables in a meaningful order

b) For snowfall, the most commonly observed values is 0. Because the dataset covers the whole year, but snow is usually only possible in winter, most observations do not have snow.


## Problem 2



## Problem 3

