---
title: "p8105_hw3_jz3902"
author: "Jinghan Zhao"
date: "2024-10-11"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)
knitr::opts_chunk$set(fig.width=10, fig.height=5)
```

## Problem 1

### Short description

```{r import_noaa}
library(p8105.datasets)
data("ny_noaa")

ny_noaa
```

There are `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns in the dataset.

```{r missing_noaa}
colMeans(is.na(ny_noaa))
```

We have missing values in `prcp`, `snow`, `snwd`, `tmax` and `tmin` columns. Especially, there over 40% data is missing in `tmax` and `tmin`, which is a rather huge proportion. 

This issue may result in a large amount of observations being ignored during data analysis, making the analysis results incomplete and unable to reflect the true situation.


### Data cleaning

```{r data_cleaning_noaa}
ny_noaa_clean = 
  ny_noaa %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    prcp = prcp / 10,
    tmax = tmax / 10,
    tmin = tmin / 10
  ) %>% 
  relocate(id, date, year, month, day)

ny_noaa_clean %>% 
  count(snow) %>% 
  arrange(desc(n)) %>% 
  head()
```

a) Data cleaning steps:

* Create variables: `year`, `month` and `day`

* Arrange the units of variables: `prcp`(mm), `snow`(mm), `snwd`(mm), `tmax`(degrees C), `tmin`(degrees C)

* Organize variables in a meaningful order

b) For snowfall, the most commonly observed values is 0. Because the dataset covers the whole year, but snow usually occurs in winter, most observations do not have snow.


### Average tmax in Jan and Jul

```{r avg_tmax, message = FALSE}
month.labs = c("January", "July")
names(month.labs) = c(1, 7)

ny_noaa_clean %>% 
  filter(month %in% c(1, 7)) %>%
  group_by(id, year, month) %>% 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>% 
  drop_na(avg_tmax) %>% 
  ggplot(aes(x = year, y = avg_tmax, color = id)) +
  geom_point() +
  geom_line() +
  facet_grid(. ~ month, 
             labeller = labeller(month = month.labs)) +
  labs(
    title = "Average Maximum Temperature in Each Station",
    x = "Year",
    y = "Average Max Temp (°C)",
    caption = "The average max temperature in January and in July in each station across years.") +
  viridis::scale_color_viridis(discrete = TRUE,
                               option = "viridis") +
  theme(legend.position = "none")
```

* Most of the time, the average maximum temperatures at different stations have similar trends: they rise or fall at the same time. 

* According to the graph, outliers occurred in July between 1985-1990. To find out which years they are:

```{r avg_tmax_outlier}
ny_noaa_clean %>% 
  filter(month == 7) %>%
  group_by(id, year, month) %>% 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) %>% 
  drop_na(avg_tmax) %>% 
  arrange(avg_tmax) %>% 
  head()
```

* Station USC00308962 have extreme low average temperatures (14.0 degrees C) in January, 1988.


### tmax vs tmin + snowfall

```{r tmax_tmin_snowfall}
hex_tmax_tmin = 
  ny_noaa_clean %>% 
  drop_na(tmin, tmax) %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(
    title = "Hexbin Plot: tmax vs tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8, angle = 45))
  

box_snowfall = 
  ny_noaa_clean %>% 
  select(year, snow) %>% 
  filter(snow > 0 & snow < 100) %>% 
  ggplot(aes(x = factor(year), y = snow)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Distribution of Snowfall (between 0-100 mm)",
       x = "Year",
       y = "Snowfall (mm)") +
  theme(axis.text.x = element_text(size = 8,
                                   angle = 45, 
                                   hjust = 1))

hex_tmax_tmin + box_snowfall
```




## Problem 2



## Problem 3

